Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-07 talk/pepm2011/paper.tex
rename operation to "get" and "set".

\begin{figure}
\texttt{
\begin{tabular}{l} 
\# arguments to the trace: $p_{0}$, $p_{1}$ \\
\# inside f: res.add(y) \\
guard\_class($p_{1}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{2}$ = getfield($p_{1}$, intval) \\
~~~~guard\_class($p_{0}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{3}$ = getfield($p_{0}$, intval) \\
~~~~~~~~$i_{4}$ = int\_add($i_{2}$, $i_{3}$) \\
~~~~~~~~$p_{5}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~setfield($p_{5}$, intval, $i_{4}$) \\
\# inside f: BoxedInteger(-100)  \\
$p_{6}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~setfield($p_{6}$, intval, -100) \\
~\\
\# inside f: .add(BoxedInteger(-100)) \\
guard\_class($p_{5}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{7}$ = getfield($p_{5}$, intval) \\
~~~~guard\_class($p_{6}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{8}$ = getfield($p_{6}$, intval) \\
~~~~~~~~$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
~~~~~~~~$p_{10}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~setfield($p_{10}$, intval, $i_{9}$) \\
~\\
\# inside f: BoxedInteger(-1) \\
$p_{11}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~setfield($p_{11}$, intval, -1) \\
~\\
\# inside f: y.add(BoxedInteger(-1)) \\
guard\_class($p_{0}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{12}$ = getfield($p_{0}$, intval) \\
~~~~guard\_class($p_{11}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{13}$ = getfield($p_{11}$, intval) \\
~~~~~~~~$i_{14}$ = int\_add($i_{12}$, $i_{13}$) \\
~~~~~~~~$p_{15}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~setfield($p_{15}$, intval, $i_{14}$) \\
~\\
\# inside f: y.is\_positive() \\
guard\_class($p_{15}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.is\_positive \\
~~~~$i_{16}$ = getfield($p_{15}$, intval) \\
~~~~$i_{17}$ = int\_gt($i_{16}$, 0) \\
\# inside f \\
guard\_true($i_{17}$) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}
\label{fig:unopt-trace}
\caption{Unoptimized Trace for the Simple Object Model}
\end{figure}

If the function is executed using the tracing JIT, with \texttt{y} being a
\texttt{BoxedInteger}, the produced trace looks like
Figure~\ref{fig:unopt-trace}. The operations in the trace are shown indented to
correspond to the stack level of the function that contains the traced
operation. The trace also shows the inefficiencies of \texttt{f} clearly, if one
looks at the number of \texttt{new} (corresponding to object creation),
\texttt{set/getfield} (corresponding to attribute reads/writes) and
\texttt{guard\_class} operations (corresponding to method calls).

@@

The partial evaluation works by walking the trace from beginning to end.
Whenever a \texttt{new} operation is seen, the operation is removed and a static
object is constructed and associated with the variable that would have stored
the result of \texttt{new}. The static object describes the shape of the
original object, \eg where the values that would be stored in the fields of the
allocated object come from, as well as the type of the object. Whenever the
optimizer sees a \texttt{setfield} that writes into such an object, that shape
description is updated and the operation can be removed, which means that the
operation was done at partial evaluation time. When the optimizer encounters a
\texttt{getfield} from such an object, the result is read from the shape
description, and the operation is also removed. Equivalently, a
\texttt{guard\_class} on a variable that has a shape description can be removed
as well, because the shape description stores the type.

@@

\texttt{
\begin{tabular}{l} 
$p_{5}$ = new(BoxedInteger) \\
setfield($p_{5}$, intval, $i_{4}$) \\
$p_{6}$ = new(BoxedInteger) \\
setfield($p_{6}$, intval, -100) \\
\end{tabular}
}

@@

\texttt{
\begin{tabular}{l} 
guard\_class($p_{5}$, BoxedInteger) \\
$i_{7}$ = getfield($p_{5}$, intval) \\
\# inside BoxedInteger.add \\
guard\_class($p_{6}$, BoxedInteger) \\
\# inside BoxedInteger.add\_\_int \\
$i_{8}$ = getfield($p_{6}$, intval) \\
$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
\end{tabular}
}

The \texttt{guard\_class} operations can be removed, because the classes of $p_{5}$ and
$p_{6}$ are known to be \texttt{BoxedInteger}. The \texttt{getfield} operations can be removed
and $i_{7}$ and $i_{8}$ are just replaced by $i_{4}$ and -100. Thus the only
remaining operation in the optimized trace would be:

@@

\texttt{
\begin{tabular}{l} 
$p_{15}$ = new(BoxedInteger) \\
setfield($p_{15}$, intval, $i_{14}$) \\
$p_{10}$ = new(BoxedInteger) \\
setfield($p_{10}$, intval, $i_{9}$) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}

Note how the operations for creating these two instances have been moved down the
trace. It looks like for these operations we actually didn't win much, because
the objects are still allocated at the end. However, the optimization was still
worthwhile even in this case, because some operations that have been performed
on the lifted static objects have been removed (some \texttt{getfield} operations
and \texttt{guard\_class} operations).

@@

XXX think of a way to format the operations to make them look distinct from
normal text

@@

Traces are lists of operations. The operations considered here are new (to make
a new object), get (to read a field out of an object), set (to write a field
into an object) and guard (to check the type of an object). The values of all
variables are locations (i.e.~pointers). Locations are mapped to objects, which
are represented by triples of a type $T$, and two locations that represent the
fields of the object. When a new object is created, the fields are initialized
to null, but we require that they are immediately initialized to a real
location, otherwise the trace is malformed.

@@

The get operation reads a field $F$ out of an object and adds the result
variable to the environment, mapping to the read location. The heap is
unchanged.

The set operation changes field $F$ of an object stored at the location that
variable $v$ maps to. The new value of the field is the location in variable
$u$. The environment is unchanged.

The guard operation is used to check whether the object stored at the location
that variable $v$ maps to is of type $T$. If that is the case, then execution
continues without changing heap and environment. Otherwise, execution is
stopped.

@@

If the argument $u$ of a get operation is mapped to something in the static
heap, the get can be performed at optimization time. Otherwise, the get
operation needs to be emitted.

If the first argument $v$ to a set operation is mapped to something in the
static heap, then the set can performed at optimization time and the static heap
is updated. Otherwise the set operation needs to be emitted. This needs to be
done carefully, because the new value for the field stored in the variable $u$
could itself be static, in which case it needs to be lifted first.

I a guard is performed on a variable that is in the static heap, the type check
can be performed at optimization time, which means the operation can be removed
if the types match. If the type check fails statically or if the object is not
in the static heap, the guard is put into the residual trace. This also needs to
lift the variable on which the guard is performed.

>>>>

\begin{figure}
\texttt{
\begin{tabular}{l} 
\# arguments to the trace: $p_{0}$, $p_{1}$ \\
\# inside f: res.add(y) \\
guard\_class($p_{1}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{2}$ = get($p_{1}$, intval) \\
~~~~guard\_class($p_{0}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{3}$ = get($p_{0}$, intval) \\
~~~~~~~~$i_{4}$ = int\_add($i_{2}$, $i_{3}$) \\
~~~~~~~~$p_{5}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{5}$, intval, $i_{4}$) \\
\# inside f: BoxedInteger(-100)  \\
$p_{6}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~set($p_{6}$, intval, -100) \\
~\\
\# inside f: .add(BoxedInteger(-100)) \\
guard\_class($p_{5}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{7}$ = get($p_{5}$, intval) \\
~~~~guard\_class($p_{6}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{8}$ = get($p_{6}$, intval) \\
~~~~~~~~$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
~~~~~~~~$p_{10}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{10}$, intval, $i_{9}$) \\
~\\
\# inside f: BoxedInteger(-1) \\
$p_{11}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~set($p_{11}$, intval, -1) \\
~\\
\# inside f: y.add(BoxedInteger(-1)) \\
guard\_class($p_{0}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{12}$ = get($p_{0}$, intval) \\
~~~~guard\_class($p_{11}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{13}$ = get($p_{11}$, intval) \\
~~~~~~~~$i_{14}$ = int\_add($i_{12}$, $i_{13}$) \\
~~~~~~~~$p_{15}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{15}$, intval, $i_{14}$) \\
~\\
\# inside f: y.is\_positive() \\
guard\_class($p_{15}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.is\_positive \\
~~~~$i_{16}$ = get($p_{15}$, intval) \\
~~~~$i_{17}$ = int\_gt($i_{16}$, 0) \\
\# inside f \\
guard\_true($i_{17}$) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}
\label{fig:unopt-trace}
\caption{Unoptimized Trace for the Simple Object Model}
\end{figure}

If the function is executed using the tracing JIT, with \texttt{y} being a
\texttt{BoxedInteger}, the produced trace looks like
Figure~\ref{fig:unopt-trace}. The operations in the trace are shown indented to
correspond to the stack level of the function that contains the traced
operation. The trace also shows the inefficiencies of \texttt{f} clearly, if one
looks at the number of \texttt{new} (corresponding to object creation),
\texttt{set/get} (corresponding to attribute reads/writes) and
\texttt{guard\_class} operations (corresponding to method calls).

@@

The partial evaluation works by walking the trace from beginning to end.
Whenever a \texttt{new} operation is seen, the operation is removed and a static
object is constructed and associated with the variable that would have stored
the result of \texttt{new}. The static object describes the shape of the
original object, \eg where the values that would be stored in the fields of the
allocated object come from, as well as the type of the object. Whenever the
optimizer sees a \texttt{set} that writes into such an object, that shape
description is updated and the operation can be removed, which means that the
operation was done at partial evaluation time. When the optimizer encounters a
\texttt{get} from such an object, the result is read from the shape
description, and the operation is also removed. Equivalently, a
\texttt{guard\_class} on a variable that has a shape description can be removed
as well, because the shape description stores the type.

@@

\texttt{
\begin{tabular}{l} 
$p_{5}$ = new(BoxedInteger) \\
set($p_{5}$, intval, $i_{4}$) \\
$p_{6}$ = new(BoxedInteger) \\
set($p_{6}$, intval, -100) \\
\end{tabular}
}

@@

\texttt{
\begin{tabular}{l} 
guard\_class($p_{5}$, BoxedInteger) \\
$i_{7}$ = get($p_{5}$, intval) \\
\# inside BoxedInteger.add \\
guard\_class($p_{6}$, BoxedInteger) \\
\# inside BoxedInteger.add\_\_int \\
$i_{8}$ = get($p_{6}$, intval) \\
$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
\end{tabular}
}

The \texttt{guard\_class} operations can be removed, because the classes of $p_{5}$ and
$p_{6}$ are known to be \texttt{BoxedInteger}. The \texttt{get} operations can be removed
and $i_{7}$ and $i_{8}$ are just replaced by $i_{4}$ and -100. Thus the only
remaining operation in the optimized trace would be:

@@

\texttt{
\begin{tabular}{l} 
$p_{15}$ = new(BoxedInteger) \\
set($p_{15}$, intval, $i_{14}$) \\
$p_{10}$ = new(BoxedInteger) \\
set($p_{10}$, intval, $i_{9}$) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}

Note how the operations for creating these two instances have been moved down the
trace. It looks like for these operations we actually didn't win much, because
the objects are still allocated at the end. However, the optimization was still
worthwhile even in this case, because some operations that have been performed
on the lifted static objects have been removed (some \texttt{get} operations
and \texttt{guard\_class} operations).

@@

@@

Traces are lists of operations. The operations considered here are \texttt{new} (to make
a new object), \texttt{get} (to read a field out of an object), \texttt{set} (to write a field
into an object) and \texttt{guard\_class} (to check the type of an object). The values of all
variables are locations (i.e.~pointers). Locations are mapped to objects, which
are represented by triples of a type $T$, and two locations that represent the
fields of the object. When a new object is created, the fields are initialized
to null, but we require that they are immediately initialized to a real
location, otherwise the trace is malformed.

@@

The \texttt{get} operation reads a field $F$ out of an object and adds the result
variable to the environment, mapping to the read location. The heap is
unchanged.

The \texttt{set} operation changes field $F$ of an object stored at the location that
variable $v$ maps to. The new value of the field is the location in variable
$u$. The environment is unchanged.

The \texttt{guard\_class} operation is used to check whether the object stored at the location
that variable $v$ maps to is of type $T$. If that is the case, then execution
continues without changing heap and environment. Otherwise, execution is
stopped.

@@

If the argument $u$ of a \texttt{get} operation is mapped to something in the static
heap, the get can be performed at optimization time. Otherwise, the \texttt{get}
operation needs to be emitted.

If the first argument $v$ to a \texttt{set} operation is mapped to something in the
static heap, then the \texttt{set} can performed at optimization time and the static heap
is updated. Otherwise the \texttt{set} operation needs to be emitted. This needs to be
done carefully, because the new value for the field stored in the variable $u$
could itself be static, in which case it needs to be lifted first.

I a \texttt{guard\_class} is performed on a variable that is in the static heap, the type check
can be performed at optimization time, which means the operation can be removed
if the types match. If the type check fails statically or if the object is not
in the static heap, the \texttt{guard\_class} is put into the residual trace. This also needs to
lift the variable on which the \texttt{guard\_class} is performed.
